---
title: "Intro to GLM"
author: "Chris McKennan"
date: "1/23/2020"
output: html_document
---

This code provides an introduction to fitting logistic and Poisson (with log link) regression models. The relevant datasets are crab.txt, poissonreg.csv and Horse.txt.

#Load data
```{r LoadData}
Data <- data.frame(read.table(file = "crab.txt", header = T, sep = "\t", stringsAsFactors = F))
Data$Y <- as.numeric(Data$satell > 0)
Data$weight <- Data$weight/1000
Data$color <- Data$color-1
Data$light <- as.numeric(Data$color < 4)
```

#GLM
Logistic regression
```{r Fit_GLM}
fit.1 <- glm(Y ~ width, family = binomial(link = "logit"), data = Data)
```

#Coefficient estimates
```{r Coefficients}
#Coefficient estimates#
beta.hat <- fit.1$coefficients

#Odds ratio associated with 1cm increase in width#
or.1cm <- exp(beta.hat[2])

#Probability a female with width=21 has a male satellite#
x.0 <- 21
prob.21 <- exp(beta.hat[1] + beta.hat[2]*x.0)/(1 + exp(beta.hat[1] + beta.hat[2]*x.0))
```

#Standard errors based on asymptotic variance
```{r SE}
#Estimate for Fisher Information#
FI <- summary(fit.1)$cov.scaled

#Standard error for estimates#
se <- sqrt(diag(FI))
```

#Wald hypothesis testing and CIs
```{r Wald}
#H_0: beta_width = 0 <=> H_0: OR = 1#
t.H0 <- beta.hat[2]/se[2]
pvalue.H0 <- 2*pnorm(-abs(t.H0))   #Very significant

#95% CI for beta#
alpha <- 0.05
CI.beta <- beta.hat[2] + se[2]*c(qnorm(alpha/2), qnorm(1-alpha/2))

#What about 95% CI for P(Y = 1 | X = 21)?#
x <- c(1,21)  #Intercept and width
se.x <- sum(x*(FI%*%x))
CI.logit <- sum(x*beta.hat) + se.x*c(qnorm(alpha/2), qnorm(1-alpha/2))
CI.prob <- exp(CI.logit)/(1+exp(CI.logit))
```

#Likelihood ratio test
Test $H_0: \beta_{width}=0$
```{r LRT}
fit.0 <- glm(Y ~ 1, family = binomial(link = "logit"), data = Data)
lr.stat <- fit.0$deviance - fit.1$deviance
pvalue.H0.lr <- pchisq(q = lr.stat, df = 1, lower.tail = F)  #Also very significant, which is not surprising
```

#Pearson
```{r CrabPearson}
fit.pearson.0 <- glm(Y ~ color, family = binomial(link = "logit"), data = Data)

Observed <- t(sapply(1:length(unique(Data$color)), function(i){ind.i <- Data$color==i; c(sum(Data$Y[ind.i]),sum(ind.i)-sum(Data$Y[ind.i]))}))
rownames(Observed) <- c("1", "2", "3", "4"); colnames(Observed) <- c("Y=1", "Y=0")

predicted <- exp(predict(fit.pearson.0))/(1+exp(predict(fit.pearson.0)))
Expected <- t(sapply(1:length(unique(Data$color)), function(i){ind.i <- Data$color==i; c(sum(predicted[ind.i]),sum(ind.i)-sum(predicted[ind.i]))}))
rownames(Expected) <- c("1", "2", "3", "4"); colnames(Expected) <- c("Y=1", "Y=0")

Pearson.X2 <- sum((Observed-Expected)^2/Expected)
pchisq(q = Pearson.X2, df = 4-2, lower.tail = F)
```

#Hosmer-Lemeshow
```{r Hosmer-Lemeshow}
library(ResourceSelection)
fit.HL <- glm(Y ~ color, family = binomial(link = "logit"), data = Data)
HL <- hoslem.test(fit.HL$y, fitted(fit.HL), g=10)
```

#Ordinal data
```{r OrdinalData}
Data$mates <- rep(0, nrow(Data))
Data$mates[Data$satell>=1 & Data$satell<=5] <- 1
Data$mates[Data$satell>5] <- 2
Data$mates <- factor(Data$mates, levels = c(0,1,2), labels = c("0","1","2"), ordered = T)
fit.ordinal <- MASS::polr(mates ~ weight + as.factor(color), data = Data)
```
Fitted model is $logit\left\lbrace \hat{P}(Y \leq 0 \mid Weight,Color=j)\right\rbrace = 1.59 + 1.22Weight + \hat{\beta}_j$ and $logit\left\lbrace \hat{P}(Y \leq 1 \mid Weight,Color=j)\right\rbrace = 3.95 + 1.22Weight + \hat{\beta}_j$. The fitted model indicates that heavier female crabs tend to have more mates.

#Poisson regression with crab data
```{r PoiCrab}
fit.crab.pois <- glm(satell ~ weight + as.factor(color), family = poisson(link="log"), data = Data)
mean.crab <- exp(predict(fit.crab.pois))
Dispersion <- sum( (Data$satell-mean.crab)^2/mean.crab )/(nrow(Data)-ncol(model.matrix(fit.crab.pois)))

#Data look over dispersed. Let's modify the dispersion parameter and see how inference changes
summary(fit.crab.pois)$coefficients
summary(fit.crab.pois, dispersion=Dispersion)$coefficients
```


#Let's switch to Poisson regression

##The data
- We have data from seniors from two high schools (159 from one school and 157 from the other).
- We recorded the number of days they were absent during their junior year as well as their
sex and their scores on standardized tests for math and language.
- We want to create a model that will explain the missing of school.
- Maybe do some prediction
```{r LoadData}
Data <- data.frame(read.csv(file = "poissonreg.csv", header = T, stringsAsFactors = F))
```

##Fit the model
```{r FitModel}
fit.poisson <- glm(daysabs ~ math + langarts + male, data=Data, family = poisson(link="log"))

summary(fit.poisson)
```
Interpretation?

###Get residuals
```{r ModelDiagnostics}
Pearson.resid <- residuals.glm(fit.poisson, type = "pearson")
Deviance.resid <- residuals.glm(fit.poisson, type = "deviance")
```

##New rate data example
- age in months
- time in days
```{r LoadData_horse}
Data <- data.frame(read.table(file ="Horse.txt", header = T, stringsAsFactors = F, sep="\t"))
```

###Fit model
```{r FitHorseModel}
fit.horse <- glm(infec ~ age, family = poisson(link="log"), offset = log(time), data=Data)
summary(fit.horse)
```

###Dispersion
```{r Dispersion}
Mean.fit <- exp(predict(fit.horse))
Dispersion <- sum( (Data$infec - Mean.fit)^2/Mean.fit )/(nrow(Data) - 2)
```