---
title: "PCA examples"
author: "Chris McKennan"
date: "4/9/2020"
output: html_document
---

Some simulation examples to understand PCA

#Simulate Data

```{r SimulationParameters}
p <- 10^4
n <- 10^2
K <- 10
#Lambda <- rep(1, length=K)   #Eigenvalues, all of which are large
Lambda <- exp(seq(log(1),log(0.5/n),length=K))   #Eigenvalues, some of which are small
```

```{r SimulatedData}
SimData <- function(p, n, Lambda) {
  out <- list()
  K <- length(Lambda)
  out$C <- matrix(rnorm(n*K), nrow=n, ncol=K)
  out$L <- matrix(NA,nrow=p,ncol=K)
  for (k in 1:K) {
    out$L[,k] <- rnorm(p)*sqrt(Lambda[k])
  }
  mu <- t(scale(x = t(out$L%*%t(out$C)),center = T, scale = F))
  out$C.true <- sqrt(n)*svd(mu)$v[,1:K]
  out$Y <- out$L%*%t(out$C) + matrix(rnorm(n*p),nrow=p,ncol=n)
  out$C.stand <- scale(x = out$C, center = T, scale = F)
  return(out)
}

out.sim <- SimData(p,n,Lambda)
```

#Analyze data
```{r AnalyzeData}
svd.sim <- svd(t( scale(x = t(out.sim$Y), center = T, scale = F) ), nu = K, nv = K)   #Assume K is known
C.hat <- sqrt(n)*svd.sim$v[,1:K]
sapply(X = 1:K, function(k){ abs(cor(C.hat[,k],out.sim$C.true[,k])) })  #Correlation between individual vectors

#Correlation between subspaces#
cos.theta <- svd(t(svd(C.hat)$u)%*%svd(out.sim$C)$u)$d[K]
```

#Choosing K with a scree plot in simulated data
```{r ScreePlot_SimulatedData}
#From simulated data#
Y.sim <- t(scale(t(out.sim$Y),center = T, scale = F))   #mean-center the rows of Y
eigs <- eigen(1/p*t(Y.sim)%*%Y.sim)$values
plot(eigs/sum(eigs), xlab="Eigenvalue index", ylab="% variance explained")
points(K, eigs[K]/sum(eigs), pch=4, col="red")
lines(1:ncol(out.sim$Y),eigs/sum(eigs))
legend("topright", pch=4, col="red", legend="True K=10")
```

#Choosing K with a scree plot in real data
Real gene expression data
Expression of $p=12,317$ genes in $n=217$ samples.
These data cannot be shared...See slides for the scree plot output.
```{r ScreePlot_RealData}
eigs.real <- eigen(1/nrow(Y.real)*t(Y.real)%*%Y.real)$values
plot(eigs.real/sum(eigs.real), xlab="Eigenvalue index", ylab="% variance explained", main="Real gene expression data")
lines(1:ncol(Y.real),eigs.real/sum(eigs.real))
```

#Choosing K with Bai and Ng
```{r BaiNg}
Bai.Ng.K <- function(Y, K.max=NULL, Cov=NULL) {
  if (!is.null(Cov)) {
    Y <- Y%*%Compute.Q(Cov)
  }
  n <- ncol(Y)
  p <- nrow(Y)
  if (is.null(K.max)) {3/4*ncol(Y)}
  svd.Y <- svd(x = Y)$d^2
  out.log <- unlist(lapply(X = 0:(K.max-1), function(k){ log(sum(svd.Y[(k+1):n])/(n*p)) })) + (1:K.max)*(n+p)/(n*p)*log((n*p)/(n+p))
  out.reg <- unlist(lapply(X = 0:(K.max-1), function(k){ sum(svd.Y[(k+1):n])/(n*p) })) + (1:K.max)*(n+p)/(n*p)*log((n*p)/(n+p))
  return(list(obj.reg=out.reg, obj.log=out.log, K.hat.log=which.min(out.log), K.hat.reg=which.min(out.reg)))
}

out.BaiNg <- Bai.Ng.K(Y = out.sim$Y, K.max = 30, Cov=rep(1,ncol(out.sim$Y)))
out.BaiNg$K.hat.reg   #R(k) is defined on the slides, P(n,p) = (n+p)/(n*p)*log((n*p)/(n+p))
out.BaiNg$K.hat.log   #R(k) is defined as the log(expression on slides), P(n,p) is the same.
```
Even in idealized settings, Bai & Ng (2002) fails when some of the eigenvalues are small

#Parallel analysis
```{r ParallelAnalysis}
DistnE <- function(Y, Cov=NULL, B=20) {
  if (!is.null(Cov)) {Y <- Y%*%Compute.Q(Cov)}
  n <- ncol(Y); p <- nrow(Y)
  out  <- matrix(NA, nrow=20, ncol=B)
  for (b in 1:B) {
    E.tilde <- t(apply(X = Y, MARGIN = 1, function(y){y[order(runif(length(y)))]}))
    out[,b] <- svd(E.tilde, nu = 20, nv = 20)$d[1:20]
  }
  return(out)
}

out.distE <- DistnE(out.sim$Y)
svd.y <- svd(out.sim$Y)
k <- 8
hist(out.distE[k,], main=paste0("k = ",k), xlab="Singular values of E.tilde", xlim=c(min(c(out.distE[k,],svd.y$d[k])),max(c(out.distE[k,],svd.y$d[k]))))
abline(v=svd.y$d[k],col="red")
```